The Inconsistency in G\"odel's Ontological Argument: \\ A Success Story for AI in Metaphysics

### 2016-01-31 Papers/InProgress/IJCAI/submission.tex
Improve section 3
Improve section 3.2

For a computational analysis of G\"odel's ontological argument, the embedding of HOMLs such as \textbf{K},
\textbf{KB} and \textbf{S5} with various domain conditions (possibilist and actualist quantification)
is required. This idea has been successfully employed in related work
\cite{C40}. The embedding of HOML is in fact
straightforward. Formulas in HOML are \emph{lifted}, i.e., converted into predicates
over worlds, which are themselves explicitly represented as
terms. The logical constants of HOML are translated to HOL terms in such a way that, for instance, 
%$\neg \varphi$, $\varphi\vee\psi$
$\Box \varphi$ and $\Diamond \varphi$ (relative to a current world
$w_o$) are mapped, respectively, to the HOL formulas
$\forall w. (r w_0 w) \imp (\varphi w)$ and
$\exists w. (r w_0 w) \wedge (\varphi w)$. This form of embedding is
precisely the well-known standard translation
\cite{DBLP:journals/logcom/Ohlbach91}, which is here intra-logically
realized --- and extended for quantifiers --- in HOL by stating a set
of equations defining the logical constants (Fig.~\ref{QML_S5U}). The
resulting object logic is the HOML \textbf{K} with rigid terms and constant
domains (possibilist quantifiers). Other logics (e.g. \textbf{KB}, \textbf{S5}) can be
embedded by adding axioms that restrict the accessibility relation
$r$. Varying domains and actualist quantifiers can be simulated by
means of an existence predicate, which can be used to guard the
quantifiers. The embedding approach is, therefore, very flexible.

@@

Wider adoption of HOL theorem proving technology for
reasoning about and within embedded object logics, especially among
non-expert users, is still hindered by the gap between the syntax used
by people, when they write logical formulas with pen and paper, and
the syntax used by HOL theorem provers. Even when the syntax of the
underlying higher-order system is elegant (as is the case in
Isabelle/HOL), the embedding of HOML into HOL may easily expose
details of HOL that may be uncommon to the user, disturbing his/her
experience while using the system. To illustrate this point,
Fig.~\ref{UglyEssence} shows how the definition of essence looked like
in previous work \cite{J28}, where advanced syntax-sugaring features
were not used. It looks notably higher-order, and its style differs
significantly from the common style seen in works on modal logics and
the ontological argument. The following specific issues can be
enumerated:
\begin{enumerate}
\item $\lambda$-abstractions, a typical higher-order feature, appear explicitly in places where they didn't need to in a pure HOML formulation (cf. G\"odel's manuscript, Fig.~\ref{GoedelScript}).
\item Quantifiers appear as higher-order defined constants, and not as binders. This forces the user to read (and write) formulas of the form $\forall (\lambda x. A(x))$ instead of the more common $\forall x. A(x)$.
\item The lifted modal connectives are represented by prefixing the
  letter ``m'' (e.g. $m\wedge$ and $m\imp$). The prefix disturbs the
  user, as it constantly reminds him/her that there is something
  unusual about the modal connectives.
\item Higher-order parenthesis conventions for the application of a predicate to a term are used. 
Instead of reading $\psi(y)$, as he would expect, he has to read $(\psi ~ y)$. Outside niche areas in computer science, the former syntax is more widely known than the latter.
\end{enumerate}

@@

In the embedding presented here, in Fig.~\ref{QML_S5U}, advanced
syntax-sugaring effects provided by Isabelle were used to prevent
issues as enumerated above. The possibility to define boldface connectives allows us to drop the prefix; ``\texttt{binder}'' annotations enable modal quantifiers to be used in the standard binding way and reduce the need for explicit lambda abstractions; and a careful choice of priorities for infix connectives gives the parenthesis conventions that are more familiar to the user. As desired, the definition of essence in Fig.~\ref{Scott_S5U} is undeniably more immediately recognizable and comprehensible than the definition in Fig.~\ref{UglyEssence}. The embedding technique is now completely transparent to the user.

We hope that the syntax improvements described here will render the computer-assisted analysis of ontological arguments (as well as future works utilizing the embedding of modal logics into HOL) accessible to a wider audience and encourage philosophers to adopt logic-based artificial intelligence tools in order to conduct similar experiments in computational metaphysics.

>>>>

For a computational analysis of G\"odel's ontological argument, the embedding of HOMLs such as \textbf{K},
\textbf{KB} and \textbf{S5} with various domain conditions (possibilist and actualist quantification)
is required. This idea has been successfully employed in related work
\cite{C40}. The embedding of HOML is in fact
straightforward. Formulas in HOML are \emph{lifted}, i.e., converted into predicates
over worlds, which are themselves explicitly represented as
terms. The logical constants of HOML are translated to HOL terms in such a way that, for instance, 
%$\neg \varphi$, $\varphi\vee\psi$
$\Box \varphi$ and $\Diamond \varphi$ (relative to a current world
$w_o$) are mapped, respectively, to the HOL formulas
$\forall w. (r w_0 w) \imp (\varphi w)$ and
$\exists w. (r w_0 w) \wedge (\varphi w)$. This form of embedding is
precisely the well-known standard translation
\cite{DBLP:journals/logcom/Ohlbach91}, which is here intra-logically
realized --- and extended for quantifiers --- in HOL by stating a set
of equations defining the logical constants (Fig.~\ref{QML_S5U}). The
resulting object logic is the HOML \textbf{K} with rigid terms and constant
domains (possibilist quantifiers). Other logics (e.g. \textbf{KB}, \textbf{S5}) can be
embedded by adding axioms that restrict the accessibility relation
$r$. Varying domains and actualist quantifiers can be simulated by using
 an existence predicate to guard the
quantifiers. The embedding approach is, therefore, very flexible.

@@

Wider adoption of HOL theorem proving technology for
reasoning about and within embedded object logics, especially among
non-expert users, is still hindered by the gap between the syntax used
by people, when they write logical formulas with pen and paper, and
the syntax used by HOL theorem provers. Even when the syntax of the
underlying higher-order system is elegant (as is the case in
Isabelle/HOL), the embedding of HOML into HOL may easily expose
details of HOL that may be uncommon to the user, disturbing his/her
experience while using the system. To illustrate this point,
Fig.~\ref{UglyEssence} shows how the definition of essence looked like
in previous work \cite{J28}, where advanced syntax-sugaring features
were not used. It looks notably higher-order, and its style differs
significantly from the common style seen in works on modal logics and
the ontological argument. The following specific issues can be
enumerated:
\begin{enumerate}
\item $\lambda$-abstractions, which are typically a HOL feature, appear explicitly in places where they didn't need to in a pure HOML formulation (cf. G\"odel's manuscript, Fig.~\ref{GoedelScript}).
\item Quantifiers appear as higher-order defined constants, and not as binders. This forces the user to read (and write) formulas of the form $\forall (\lambda x. A(x))$ instead of the more common $\forall x. A(x)$.
\item The lifted modal connectives are represented by prefixing the
  letter ``m'' (e.g. $m\wedge$ and $m\imp$). The prefix disturbs the
  user, as it constantly reminds him/her that there is something
  unusual about the modal connectives.
\item Higher-order parenthesis conventions for the application of a predicate to a term are used. 
Instead of reading $\psi(y)$, as he/she would expect, he/she has to read $(\psi ~ y)$. Outside niche areas in computer science, the former syntax is more widely known than the latter.
\end{enumerate}

@@

In the embedding presented here, in Fig.~\ref{QML_S5U}, advanced
syntax-sugaring effects provided by Isabelle were used to prevent
issues as those enumerated above. The possibility to define boldface connectives allows us to drop the prefix; ``\texttt{binder}'' annotations enable modal quantifiers to be used in the standard binding way and reduce the need for explicit lambda abstractions; and a careful choice of priorities for infix connectives gives the parenthesis conventions that are more familiar to the user. As desired, the definition of essence in Fig.~\ref{Scott_S5U} is undeniably more immediately recognizable and comprehensible than the definition in Fig.~\ref{UglyEssence}. The embedding technique is now completely transparent to the user.

The syntax improvements described here render the computer-assisted analysis of ontological arguments accessible to a wider audience and ease the adoption of logic-based artificial intelligence tools by philosophers interested in topics where modal logic reasoning is required.
